---
title: "Getting Started with pgxR"
author: "Qualia Hooker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pgxR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 10,
  fig.height = 6,
  warning   = FALSE,
  message   = FALSE
)
```

## Overview

`pgxR` is a pharmacogenomics toolkit for fitting dose-response curves,
extracting sensitivity metrics (IC50, AUC, Hill slope), and visualizing
responses stratified by ancestry and genomic feature status.

The typical workflow is:
```
load_data() → combine_reps() → fit_all() → predict_drc()
                                          ↓
                                summarize_drc()  ← IC50, AUC, Hill slope
                                          ↓
                                plot_*()         ← visualization
                                          ↓
                                run_aov()        ← statistics
```

---

## 1. Load and preprocess data

`pgxR` ships with a synthetic paclitaxel dose-response dataset simulating
8 lung cancer cell lines across 4 ancestry groups (AFR, EUR, EAS, AMR) and
4 genomic feature groups (NF1-del, KRAS-mut, TP53-mut, WT).
```{r load}
library(pgxR)

data(pgxr_example)
head(pgxr_example)
```

Aggregate replicates to per-cell-line means:
```{r aggregate}
agg <- combine_reps(pgxr_example)
head(agg)
```

---

## 2. Fit dose-response models

`fit_all()` fits a 4-parameter log-logistic model (LL.4) to each cell line:
```{r fit, message = FALSE, warning = FALSE}
fits <- fit_all(agg, unique(agg$cell_line))
```

---

## 3. Generate predictions
```{r predict}
preds <- predict_drc(fits, agg)
```

---

## 4. Extract pharmacological metrics

`summarize_drc()` returns IC50, AUC, and Hill slope in a single tidy table:
```{r metrics}
metrics <- summarize_drc(fits, agg, preds)
metrics
```

IC50 and AUC capture complementary information:

- **IC50** — dose at 50% effect, undefined when curves don't reach 50%
- **AUC** — integrates the full curve, always defined, higher = more sensitive
- **Hill slope** — steepness of the transition, larger absolute value = sharper switch

---

## 5. Visualize dose-response curves

### All cell lines
```{r plot-drc, fig.cap = "Dose-response curves colored by cell line."}
plot_drc(agg, preds, title = "Paclitaxel Response")
```

### Stratified by ancestry and feature
```{r plot-ancestry, fig.cap = "DRC faceted by feature, colored by ancestry."}
plot_drc_anc(agg, preds,
             title   = "Paclitaxel Response by Ancestry",
             x_label = "Dose [nM]")
```

---

## 6. Visualize IC50 estimates
```{r plot-dot, fig.cap = "IC50 point estimates with 95% confidence intervals."}
ic50 <- extract_all(fits, agg)
plot_dot(ic50, title = "Paclitaxel IC50", units = "nM")
```
```{r plot-feature, fig.cap = "IC50 faceted by genomic feature."}
plot_feature(ic50, title = "Paclitaxel IC50 by Feature", units = "nM")
```
```{r plot-anc, fig.cap = "IC50 colored by ancestry."}
plot_anc(ic50, title = "Paclitaxel IC50 by Ancestry", units = "nM")
```

---

## 7. Visualize AUC
```{r plot-auc, fig.cap = "Normalized AUC per cell line colored by ancestry."}
auc <- calc_auc(preds)
plot_auc(auc, title = "Paclitaxel AUC")
```

---

## 8. Group comparisons with p-values
```{r plot-sensitivity, fig.width = 7, fig.cap = "IC50 by ancestry with Wilcoxon p-values."}
plot_sensitivity(metrics,
                 group_col = "ancestry",
                 metric    = "IC50",
                 units     = "nM",
                 title     = "Paclitaxel IC50 by Ancestry")
```

---

## 9. Statistical testing

### ANOVA across ancestry groups
```{r stats-anova}
res <- run_aov(metrics, group_col = "ancestry", metric = "IC50")
```

### ANOVA across feature groups
```{r stats-feature}
res_feature <- run_aov(metrics, group_col = "feature", metric = "AUC")
```

---

## 10. Working with your own data

If your CSV uses different column names or your drug is measured in µM,
use `col_map` and the `units` argument:
```{r own-data, eval = FALSE}
# load with column mapping
raw <- load_data("my_data.csv",
                 col_map = list(feature = "translocation"))

# standardize ancestry to 1KGP/HGDP abbreviations
raw$ancestry <- standardize_ancestry(as.character(raw$ancestry))

# aggregate — specify replicate column if present
agg     <- combine_reps(raw, replicate_col = "experiment")
fits    <- fit_all(agg, unique(agg$cell_line))
preds   <- predict_drc(fits, agg)
metrics <- summarize_drc(fits, agg, preds)

# use units = "\u00b5M" for micromolar data
plot_dot(metrics, units = "\u00b5M", title = "My Drug IC50")
plot_sensitivity(metrics,
                 group_col = "ancestry",
                 units     = "\u00b5M",
                 title     = "IC50 by Ancestry")
```

---

## Session info
```{r session}
sessionInfo()
```